---
title: "Getting Started with NNS: Correlation and Dependence"
author: "Fred Viole"
output: rmarkdown::html_vignette
vignette: >
  %\VignetteIndexEntry{Getting Started with NNS: Correlation and Dependence}
  %\VignetteEngine{knitr::rmarkdown}
  \usepackage[utf8]{inputenc}
---

```{r setup, include=FALSE, message=FALSE}
knitr::opts_chunk$set(echo = TRUE)
library(NNS)
library(data.table)
data.table::setDTthreads(2L)
options(mc.cores = 1)
Sys.setenv("OMP_THREAD_LIMIT" = 2)
```

```{r setup2,message=FALSE,warning = FALSE}
library(NNS)
library(data.table)
require(knitr)
require(rgl)
```

# Correlation and Dependence
The limitations of linear correlation are well known.  Often one uses correlation, when dependence is the intended measure for defining the relationship between variables.  NNS dependence **`NNS.dep`** is a signal:noise measure robust to nonlinear signals.

Below are some examples comparing NNS correlation **`NNS.cor`** and **`NNS.dep`** with the standard Pearson's correlation coefficient `cor`.

## Linear Equivalence
Note the fact that all observations occupy the co-partial moment quadrants.
```{r linear,fig.width=5,fig.height=3,fig.align = "center"}
x = seq(0, 3, .01) ; y = 2 * x
```

```{r linear1,fig.width=5,fig.height=3,fig.align = "center", results='hide', echo=FALSE}
NNS.part(x, y, Voronoi = TRUE, order = 3)
```

```{r res1}
cor(x, y)
NNS.dep(x, y)
```

## Nonlinear Relationship
Note the fact that all observations occupy the co-partial moment quadrants.
```{r nonlinear,fig.width=5,fig.height=3,fig.align = "center", results='hide'}
x = seq(0, 3, .01) ; y = x ^ 10
```

```{r nonlinear1,fig.width=5,fig.height=3,fig.align = "center", results='hide', echo=FALSE}
NNS.part(x, y, Voronoi = TRUE, order = 3)
```

```{r res2a}
cor(x, y)
NNS.dep(x, y)
```


## Cyclic Relationship
Even the difficult inflection points, which span both the co- and divergent partial moment quadrants, are properly compensated for in **`NNS.dep`**.
```{r nonlinear_sin,fig.width=5,fig.height=3,fig.align = "center", results='hide'}
x = seq(0, 12*pi, pi/100) ; y = sin(x)
```

```{r nonlinear1_sin,fig.width=5,fig.height=3,fig.align = "center", results='hide', echo=FALSE}
NNS.part(x, y, Voronoi = TRUE, order = 3, obs.req = 0)
```

```{r res2_sin}
cor(x, y)
NNS.dep(x, y)
```

## Dependence
Note the fact that all observations occupy only co- or divergent partial moment quadrants for a given subquadrant.
```{r dependence,fig.width=5,fig.height=3,fig.align = "center"}
set.seed(123)
df <- data.frame(x = runif(10000, -1, 1), y = runif(10000, -1, 1))
df <- subset(df, (x ^ 2 + y ^ 2 <= 1 & x ^ 2 + y ^ 2 >= 0.95))
```

```{r circle1,fig.width=5,fig.height=3,fig.align = "center", results='hide', echo=FALSE}
NNS.part(df$x, df$y, Voronoi = TRUE, order = 3, obs.req = 0)
```

```{r res3}
NNS.dep(df$x, df$y)
```




# p-values for `NNS.dep()`
p-values and confidence intervals can be obtained from sampling random permutations of $y \rightarrow y_p$ and running **`NNS.dep(x,$y_p$)`** to compare against a null hypothesis of 0 correlation, or independence between $(x, y)$.

Simply set **`NNS.dep(..., p.value = TRUE, print.map = TRUE)`** to run 100 permutations and plot the results.

```{r permutations}
## p-values for [NNS.dep]
x <- seq(-5, 5, .1); y <- x^2 + rnorm(length(x))
```

```{r perm1,fig.width=5,fig.height=3,fig.align = "center", results='hide', echo=FALSE}
NNS.part(x, y, Voronoi = TRUE, order = 3)
```

```{r permutattions_res,fig.width=5,fig.height=3,fig.align = "center"}
NNS.dep(x, y, p.value = TRUE, print.map = TRUE)
```

# Multivariate Dependence `NNS.copula()`
These partial moment insights permit us to extend the analysis to multivariate
instances and deliver a dependence measure $(D)$ such that $D \in [0,1]$. This level of analysis is simply impossible with Pearson or other rank
based correlation methods, which are restricted to bivariate cases.

```{r multi, warning=FALSE}
set.seed(123)
x <- rnorm(1000); y <- rnorm(1000); z <- rnorm(1000)
NNS.copula(cbind(x, y, z), plot = TRUE, independence.overlay = TRUE)
```


# References
If the user is so motivated, detailed arguments and proofs are provided within the following:

* [Nonlinear Nonparametric Statistics: Using Partial Moments](https://github.com/OVVO-Financial/NNS/blob/NNS-Beta-Version/examples/index.md)

* [Nonlinear Correlation and Dependence Using NNS](https://doi.org/10.2139/ssrn.3010414)

* [Deriving Nonlinear Correlation Coefficients from Partial Moments](https://doi.org/10.2139/ssrn.2148522)

* [Beyond Correlation: Using the Elements of Variance for Conditional Means and Probabilities](https://doi.org/10.2139/ssrn.2745308)


```{r threads, echo = FALSE}
Sys.setenv("OMP_THREAD_LIMIT" = "")
```