---
title: "Getting Started with NNS: Partial Moments"
author: "Fred Viole"
output: rmarkdown::html_vignette
vignette: >
  %\VignetteIndexEntry{Getting Started with NNS: Partial Moments}
  %\VignetteEngine{knitr::rmarkdown}
  \usepackage[utf8]{inputenc}
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE)
require(NNS)
```

# Partial Moments

Why is it necessary to parse the variance with partial moments?  The additional information generated from partial moments permits a level of analysis simply not possible with traditional summary statistics.

Below are some basic equivalences demonstrating partial moments role as the elements of variance.

## Mean
```{r mean}
set.seed(123) ; x = rnorm(100) ; y = rnorm(100)

mean(x)
UPM(1, 0, x) - LPM(1, 0, x)
```

## Variance
```{r variance}
var(x)

# Sample Variance:
UPM(2, mean(x), x) + LPM(2, mean(x), x)

# Population Variance:
(UPM(2, mean(x), x) + LPM(2, mean(x), x)) * (length(x) / (length(x) - 1))

# Variance is also the co-variance of itself:
(Co.LPM(1, 1, x, x, mean(x), mean(x)) + Co.UPM(1, 1, x, x, mean(x), mean(x)) - D.LPM(1, 1, x, x, mean(x), mean(x)) - D.UPM(1, 1, x, x, mean(x), mean(x))) * (length(x) / (length(x) - 1))
```


## Standard Deviation
```{r stdev}
sd(x)
((UPM(2, mean(x), x) + LPM(2, mean(x), x)) * (length(x) / (length(x) - 1))) ^ .5
```

## Covariance
```{r covariance}
cov(x, y)
(Co.LPM(1, 1, x, y, mean(x), mean(y)) + Co.UPM(1, 1, x, y, mean(x), mean(y)) - D.LPM(1, 1, x, y, mean(x), mean(y)) - D.UPM(1, 1, x, y, mean(x), mean(y))) * (length(x) / (length(x) - 1))
```

## Covariance Elements and Covariance Matrix
```{r cov_dec}
PM.matrix(LPM.degree = 1, UPM.degree = 1,target = 'mean', variable = cbind(x, y), pop.adj = TRUE)
```

## Pearson Correlation
```{r pearson}
cor(x, y)
cov.xy = (Co.LPM(1, 1, x, y, mean(x), mean(y)) + Co.UPM(1, 1, x, y, mean(x), mean(y)) - D.LPM(1, 1, x, y, mean(x), mean(y)) - D.UPM(1, 1, x, y, mean(x), mean(y))) * (length(x) / (length(x) - 1))
sd.x = ((UPM(2, mean(x), x) + LPM(2, mean(x), x)) * (length(x) / (length(x) - 1))) ^ .5
sd.y = ((UPM(2, mean(y), y) + LPM(2, mean(y) , y)) * (length(y) / (length(y) - 1))) ^ .5
cov.xy / (sd.x * sd.y)
```

## CDFs (Discrete and Continuous)
```{r cdfs,fig.align="center",fig.width=5,fig.height=3}
P = ecdf(x)
P(0) ; P(1)
LPM(0, 0, x) ; LPM(0, 1, x)

# Vectorized targets:
LPM(0, c(0, 1), x)

plot(ecdf(x))
points(sort(x), LPM(0, sort(x), x), col = "red")
legend("left", legend = c("ecdf", "LPM.CDF"), fill = c("black", "red"), border = NA, bty = "n")

# Joint CDF:
Co.LPM(0, 0, x, y, 0, 0)

# Vectorized targets:
Co.LPM(0, 0, x, y, c(0, 1), c(0, 1))

# Continuous CDF:
plot(sort(x), LPM.ratio(1, sort(x), x), type = "l", col = "blue", lwd = 3, xlab = "x")
```


## PDFs
```{r pdfs,fig.align="center",fig.width=5,fig.height=3}
NNS.PDF(degree = 1, x)
```

## Numerical Integration
Partial moments are asymptotic area approximations of $f(x)$ akin to the familiar Trapezoidal and Simpson's rules.  More observations, more accuracy...

$$[UPM(1,0,f(x))-LPM(1,0,f(x))]\asymp\frac{[F(b)-F(a)]}{[b-a]}$$

```{r numerical integration}
x = seq(0, 1, .001) ; y = x ^ 2
UPM(1, 0, y) - LPM(1, 0, y)
```

$$0.3333=\frac{\int_{0}^{1} x^2 dx}{1-0}$$
For the total area, not just the definite integral, simply sum the partial moments:
$$[UPM(1,0,f(x))+LPM(1,0,f(x))]\asymp\left\lvert{\int_{a}^{b} f(x)dx}\right\rvert$$

## Bayes' Theorem
For example, when ascertaining the probability of an increase in $A$ given an increase in $B$, the `Co.UPM(degree.x, degree.y, x, y, target.x, target.y)` target parameters are set to `target.x = 0` and `target.y = 0` and the `UPM(degree, target, variable)` target parameter is also set to `target = 0`.

$$P(A|B)=\frac{Co.UPM(0,0,A,B,0,0)}{UPM(0,0,B)}$$

# References
If the user is so motivated, detailed arguments and proofs are provided within the following:

* [Nonlinear Nonparametric Statistics: Using Partial Moments](http://a.co/5bpHvUg)

* [Cumulative Distribution Functions and UPM/LPM Analysis](https://ssrn.com/abstract=2148482)

* [f(Newton)](https://ssrn.com/abstract=2186471)

* [Bayes' Theorem From Partial Moments](https://github.com/OVVO-Financial/NNS/blob/NNS-Beta-Version/examples/Bayes'%20Theorem%20From%20Partial%20Moments.pdf)
