---
title: "Getting Started with NNS: Correlation and Dependence"
author: "Fred Viole"
output: rmarkdown::html_vignette
vignette: >
  %\VignetteIndexEntry{Getting Started with NNS: Correlation and Dependence}
  %\VignetteEngine{knitr::rmarkdown}
  \usepackage[utf8]{inputenc}
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE)
require(NNS)
require(knitr)
require(rgl)
```

# Correlation and Dependence
The limitations of linear correlation are well known.  Often one uses correlation, when dependence is the intended measure for defining the relationship between variables.  NNS dependence `NNS.dep` is a signal:noise measure robust to nonlinear signals.

Below are some examples comparing NNS correlation `NNS.cor` and `NNS.dep` with the standard Pearson's correlation coefficient `cor`.

##Linear Equivalence
```{r linear}
x=seq(0,3,.01); y=2*x

cor(x,y)
NNS.dep(x,y,print.map = T)
```

##Nonlinear Relationship
```{r nonlinear}
x=seq(0,3,.01); y=x^10

cor(x,y)
NNS.dep(x,y,print.map = T)
```

##Dependence
```{r dependence}
set.seed(123)
df<- data.frame(x=runif(10000,-1,1),y=runif(10000,-1,1))
df<- subset(df, (x^2 + y^2 <= 1 & x^2 + y^2 >= 0.95))
NNS.dep(df$x,df$y,print.map = T)
```

#References
If the user is so motivated, detailed arguments and proofs are provided within the following:

*[Nonlinear Nonparametric Statistics: Using Partial Moments](http://a.co/5bpHvUg)

*[Deriving Nonlinear Correlation Coefficients from Partial Moments](https://ssrn.com/abstract=2148522)

*[Beyond Correlation: Using the Elements of Variance for Conditional Means and Probabilities](https://ssrn.com/abstract=2745308)

