---
title: "Getting Started with NNS: Clustering and Regression"
author: "Fred Viole"
output: rmarkdown::html_vignette
vignette: >
  %\VignetteIndexEntry{Getting Started with NNS: Clustering and Regression}
  %\VignetteEngine{knitr::rmarkdown}
  \usepackage[utf8]{inputenc}
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE)
require(NNS)
require(knitr)
require(rgl)
```

# Clustering and Regression
Below are some examples demonstrating unsupervised learning with NNS clustering and nonlinear regression using the resulting clusters.  As always, for a more thorough description and definition, please view the References.

##NNS Partitioning `NNS.part`
`NNS.part` is both a partitional and hierarchal clustering method.  NNS iteratively partitions the joint distribution into partial moment quadrants, and then assigns a quadrant identification at each partition.

`NNS.part` returns a data.table of observations along with their final quadrant identification.  It also returns the regression points, which are the quadrant means used in `NNS.reg`.
```{r linear}
x=seq(-5,5,.05); y=x^3

NNS.part(x,y,Voronoi = T)
```

### X-only Partitioning
`NNS.part` offers a partitioning based on $x$ values only, using the entire bandwidth in its regression point derivation, and shares the same limit condition as partitioning via both $x$ and $y$ values.
```{r x part}
NNS.part(x,y,Voronoi = T,type="XONLY",order=3)
```

##Clusters Used in Regression
```{r depreg,results='hide'}
for(i in 1:3){NNS.part(x,y,order=i,Voronoi = T);NNS.reg(x,y,order=i)}
```


##NNS Regression `NNS.reg`
`NNS.reg` can fit any $f(x)$, for both uni- and multivariate cases.  `NNS.reg` returns a self-evident list of values provided below.

###Univariate:
```{r nonlinear}
NNS.reg(x,y,order=4,noise.reduction = 'off')
```

###Multivariate:
```{r nonlinear multi,results='hide'}
f= function(x,y) x^3+3*y-y^3-3*x
y=x; z=expand.grid(x,y)
g=f(z[,1],z[,2])
NNS.reg(z,g,order='max')
```

###Inter/Extrapolation
`NNS.reg` can inter- or extrapolate any point of interest.  The `NNS.reg(x,y,point.est=...)` paramter permits any sized data of similar dimensions to $x$ and called specifically with `$Point.est`.
```{r iris point.est}
NNS.reg(iris[,1:4],iris[,5],point.est=iris[1:10,1:4])$Point.est
```

###NNS Dimension Reduction Regression
`NNS.reg` also provides a dimension reduction regression by including a parameter `NNS.reg(x,y,type="CLASS")`.  Reducing all regressors to a single dimension using the returned equation `$equation`.
```{r nonlinear class}
NNS.reg(iris[,1:4],iris[,5],type = "CLASS")$equation
```

####Threshold
`NNS.reg(x,y,type="CLASS",threshold=...)` offers a method of reducing regressors further by controlling the absolute value of required correlation.
```{r nonlinear class threshold}
NNS.reg(iris[,1:4],iris[,5],type = "CLASS",threshold=.35)$equation
```

and the `point.est=(...)` operates in the same manner as the full regression above, again called with `$Point.est`.
```{r }
NNS.reg(iris[,1:4],iris[,5],type = "CLASS",threshold=.35,point.est=iris[1:10,1:4])$Point.est
```


#References
If the user is so motivated, detailed arguments further examples are provided within the following:

*[Nonlinear Nonparametric Statistics: Using Partial Moments](http://a.co/5bpHvUg)

*[Deriving Nonlinear Correlation Coefficients from Partial Moments](https://ssrn.com/abstract=2148522)

*[New Nonparametric Curve-Fitting Using Partitioning, Regression and Partial Derivative Estimation](https://ssrn.com/abstract=2852570)

*[Clustering and Curve Fitting by Line Segments](https://ssrn.com/abstract=2861339)

*[Classification Using NNS Clustering Analysis](https://ssrn.com/abstract=2864711)
