---
title: "Getting Started with NNS: Correlation and Dependence"
author: "Fred Viole"
output: rmarkdown::html_vignette
vignette: >
  %\VignetteIndexEntry{Getting Started with NNS: Correlation and Dependence}
  %\VignetteEngine{knitr::rmarkdown}
  \usepackage[utf8]{inputenc}
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE)
```

```{r setup2,message=FALSE,warning = FALSE}
require(NNS)
require(knitr)
require(rgl)
require(data.table)
require(dtw)
```

# Correlation and Dependence
The limitations of linear correlation are well known.  Often one uses correlation, when dependence is the intended measure for defining the relationship between variables.  NNS dependence `NNS.dep` is a signal:noise measure robust to nonlinear signals.

Below are some examples comparing NNS correlation `NNS.cor` and `NNS.dep` with the standard Pearson's correlation coefficient `cor`.

## Linear Equivalence
Note the fact that all observations occupy the co-partial moment quadrants.
```{r linear,fig.width=5,fig.height=3,fig.align = "center"}
x = seq(0, 3, .01) ; y = 2 * x

cor(x, y)
NNS.dep(x, y, print.map = TRUE)
```

## Nonlinear Relationship
Note the fact that all observations occupy the co-partial moment quadrants.
```{r nonlinear,fig.width=5,fig.height=3,fig.align = "center"}
x=seq(0, 3, .01) ; y = x ^ 10

cor(x, y)
NNS.dep(x, y, print.map = TRUE)
```

## Dependence
Note the fact that all observations occupy only co- or divergent partial moment quadrants for a given subquadrant.
```{r dependence,fig.width=5,fig.height=3,fig.align = "center"}
set.seed(123)
df <- data.frame(x = runif(10000, -1, 1), y = runif(10000, -1, 1))
df <- subset(df, (x ^ 2 + y ^ 2 <= 1 & x ^ 2 + y ^ 2 >= 0.95))
NNS.dep(df$x, df$y, print.map = TRUE)
```


# Multi-Dimensional Dependence
These partial moment insights permit us to extend the analysis to multivariate
instances. This level of analysis is simply impossible with Pearson or other rank
based correlation methods, which are restricted to pairwise cases.

```{r multi}
set.seed(123)
x <- rnorm(1000); y <- rnorm(1000); z <- rnorm(1000)
NNS.dep.hd(cbind(x, y, z), plot = TRUE, independence.overlay = TRUE)
```

# p-values for `NNS.dep()`
p-values and confidence intervals can be obtained from sampling random permutations of $y \rightarrow y_p$ and running NNS.dep(x,$y_p$) to compare against a null hypothesis of 0 correlation, or independence between $(x, y)$.

```{r permutations}
## p-values for [NNS.dep]
x <- seq(-5, 5, .1); y <- x^2 + rnorm(length(x))

nns_cor_dep <- NNS.dep(x,abs (y), print.map = TRUE)
nns_cor_dep

## Create permutations of y
y_p <- replicate(100, sample.int(length(y)))

## Generate new correlation and dependence measures on each new permutation of y
nns.mc <- apply(y_p, 2, function(g) NNS.dep(x, y[g]))

## Store results
cors <- unlist(lapply(nns.mc, "[[", 1))
deps <- unlist(lapply(nns.mc, "[[", 2))

## View results
hist(cors)
abline(v = LPM.VaR(.975,0, cors), col = 'red')
abline(v = UPM.VaR(.975,0, cors), col = 'red')


## Left tailed correlation p-value
cor_p_value <- LPM(0, nns_cor_dep$Correlation, cors)
cor_p_value

## Right tailed correlation p-value
cor_p_value <- UPM(0, nns_cor_dep$Correlation, cors)
cor_p_value

## Confidence Intervals
## For 95th percentile VaR (both-tails) see [LPM.VaR] and [UPM.VaR]
## Lower CI
LPM.VaR(.975, 0, cors)
## Upper CI
UPM.VaR(.975, 0, cors)


hist(deps)
abline(v = LPM.VaR(.975,0, deps), col = 'red')
abline(v = UPM.VaR(.975,0, deps), col = 'red')


## Left tailed dependence p-value
dep_p_value <- LPM(0, nns_cor_dep$Dependence, deps)
dep_p_value

## Right tailed dependence p-value
dep_p_value <- UPM(0, nns_cor_dep$Dependence, deps)
dep_p_value

## Confidence Intervals
## For 95th percentile VaR (both-tails) see [LPM.VaR] and [UPM.VaR]
## Lower CI
LPM.VaR(.975, 0, deps)
## Upper CI
UPM.VaR(.975, 0, deps)
```

# References
If the user is so motivated, detailed arguments and proofs are provided within the following:

* [Nonlinear Nonparametric Statistics: Using Partial Moments](http://a.co/5bpHvUg)

* [Nonlinear Correlation and Dependence Using NNS](https://ssrn.com/abstract=3010414)

* [Deriving Nonlinear Correlation Coefficients from Partial Moments](https://ssrn.com/abstract=2148522)

* [Beyond Correlation: Using the Elements of Variance for Conditional Means and Probabilities](https://ssrn.com/abstract=2745308)

